Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-07-16 response.tex

\documentclass[wcp]{jmlr}

\usepackage{enumerate}
\usepackage{bbold}
\usepackage{booktabs}
\usepackage{natbib}
\usepackage[utf8]{inputenc}

\jmlryear{2014}
\jmlrworkshop{Neural Connectomics Workshop}

\title{Simple connectome inference from partial correlation statistics in calcium imaging}

  \author{\Name{Antonio Sutera},
   \Name{Arnaud Joly},
   \Name{Vincent François-Lavet}, \Email{a.sutera@ulg.ac.be}\\
   \Name{Zixiao Aaron Qiu},
   \Name{Gilles Louppe},
   \Name{Damien Ernst}\and\Name{Pierre Geurts}
    \\
   \addr Department of EE and CS \& GIGA-R, University of Li\`ege, Belgium}

% \author{\Name{Antonio Sutera,
%     Arnaud Joly,
%         Vincent François-Lavet,\\
%         Aaron Qiu,
%         Gilles Louppe,
%         Damien Ernst,
%         Pierre Geurts}
%   \Email{a.sutera@ulg.ac.be}\\
%   \addr Department of EE and CS \& GIGA-R, University of Li\`ege, Belgium}

% \jmlrauthors{A. Sutera et al}
% \shortauthor{A. Sutera et al.}

\begin{document}

\maketitle
\begin{center}
\vspace{-1.5cm}
\today
\end{center}

We are grateful to the reviewers for reviewing our paper and we wish to thank them for providing us with helpful and constructive criticisms.
We have modified our paper to take into account as best possible the reviewers' advices and we believe that our paper will benefit from that review.

\section*{Reviewer 1:}
\textcolor{blue}{
Better justify use of specific low-pass filters. How these filters were designed? Is this standard in the signal processing field? Will these filters apply to other problems or they need fine-tuning?}

\textbf{A}: The first low-pass filter is standard in the signal processing field while the second is designed to take into account the time. This second filter is hovewever quite classical. We add justification and motivation in the paper.\\

\noindent
\textcolor{blue}{Is there sufficient sample and power to condition on all features excluding Xi and Xj when computing partial correlation coefficient?}

\textbf{A}: It is mentioned in the paper that there are enough samples to invert the covariance matrix. We have not done a statistical analysis in the scope of this work because we only use the partial correlation to rank pairs and therefore a statistical analysis is not really needed to make that ranking. However, it would be interesting to do it in future works.

\section*{Reviewer 2:}

\textcolor{blue}{It would be necessary to have information about the computational complexity, scalability, memory requirements, and execution time for these methods.}

\textbf{A}: We obviously agree with the reviewer that such informations are interesting, especially in the context of a challenge. We add a table with some interesting figures. However, let us note that the simple method presented in the paper already gives good performances despite its light computational needs.

We also would like to mention that our code is available publicly on \url{https://github.com/asutera/kaggle-connectomics} and that some verifications, including memory requirements and execution time, of the full method are displayed on the challenge website \url{http://connectomics.chalearn.org/home/challenge-results-verification}.\\

% \textcolor{blue}{- Soundness (good experiments, good proofs):  4\\}
\noindent
\textcolor{blue}{In section 4, PC, GTE, and GENIE3 are compared with other methods. Another worthwhile experiment would be to extract features using the procedure described by the authors, do a Principal Component Analysis(PCA), and then apply PC, etc. other methods.}

\textbf{A}: We are not sure what exactly suggests the reviewer but applying PCA before PC, GTE and GENIE3 is not straighforward. Let us note that for each method we choose the preprocessing which yields the best performance.\\

\noindent
\textcolor{blue}{The authors have also used an additional relevant metric the AUPR beyond the challenge.}

\textbf{A}: We thank the reviewer to note the interest of this second metric.\\

% - Clarity: 5. 
\noindent
\textcolor{blue}{The paper, in general, is well organized. A figure/flowchart describing the workflow would be good visualization.}

\textbf{A}: Because of the constraint on the number of pages (i.e., $6$), it is not really possible to include such a figure. But we agree the reviewer that such a figure would definitely improves the paper.\\

% - Novelty/originality/insight: 4. 
\noindent
\textcolor{blue}{The paper describes preprocessing and filtering of calcium fluorescence data followed by partial correlation to identify connections.  The authors adequately describe the intuition behind the several steps of filtering they have applied.  The techniques are not new. However, this approach yielded great results.}

\textbf{A}: We wish to thank the reviewer and we are happy that he or she points out that this approach yields great results.\\

\noindent
\textcolor{blue}{“The human brain is a complex……100 billion of neurons, each connected to 7000 other neurons on average” citation needed.\\
Section 2, equation 4, symbols in equation need to be described.}
%Arnaud?\\

\textbf{A}: Introduction (Section~1) and Section~2 have been modified accordingly.\\

\noindent
\textcolor{blue}{In section 3, M principal components are used. However, the choice of M = 0.8p is not justified.}

\textbf{A}: We agree with the reviewer that this choice is not justified in the paper. We have thus included in the appendix a justication of this specific value for the parameter $M$.

%In Appendix?\\

\section*{Reviewer 3:}

\textcolor{blue}{The authors have concisely and clearly summarized their algorithm that led them to win the Connectomics Challenge. This paper is critical as a detailed explication of the winning algorithm is very useful to anyone desiring to implement this method, which will no doubt include many workshop attendees.  I have only minor corrections regarding some of the background explanations as I think clarity on the source of the problems faced by this challenge is crucial.}

\textbf{A}: We wish to thank the reviewer for his or her judicious comments about the background explanations.\\

\noindent
\textcolor{blue}{Page 2, “signal is very noisy due to light scattering”  This is not entirely correct.  The signal is noisy due to many factors including but not limited to calcium fluctuations independent of spiking activity, calcium fluctuations in nearby tissues which may be erroneously included in the region of interest due to poor region selection or poor imaging resolution, fluorescence fluctuations independent of calcium, and imaging noise in general (shot noise, dark noise,  electrical noise in the acquisition/digitisation system, etc).}

\noindent
\textcolor{blue}{Page 2, “neurons communicate through short spikes, characterized by a high frequency, while low frequencies of the signal mainly correspond to the slow decay of fluorescence”  Neurons generate electrical ‘spikes’, or action potentials, which are indeed characterized by a high frequency. However, the sharp rise in the signal observed in the fluorescent calcium indicator traces is due to the sudden influx of calcium that enters the cell during the action potential.  The decay is due to the slow decay, as stated.  But neurons communicate via neurotransmitter released by an action potential at synapses between the neurons. The ‘short spikes’ observed in the fluorescent traces therefore only very indirectly indicate whether two ‘neurons communicate’.}

\textbf{A}: We agree and we have modified Section~2 accordingly to take into account these supplementary explanations.\\

\end{document}

### 2014-07-17 article.tex
add section in appendix about the value of parameter M

\begin{abstract}
In this work, we propose a simple, but yet efficient method for the problem of
connectome inference in calcium imaging data. The proposed algorithm consists of
two steps. First, processing the raw signals to detect neural peak activities.
Second, inferring the degree of association between neurons from partial
correlation statistics. This paper summarizes the methodology that led us to
win the Connectomics Challenge, proposes a simplified version of our method and
finally discusses our results with respect to other inference methods.
\end{abstract}

@@

The human brain is a complex biological
organ made of 100 billions of neurons, each connected to 7000 other neurons on
average. Unfortunately, direct observation of the connectome, the wiring
diagram of the brain, is not yet technically feasible. Without being perfect,
calcium imaging currently allows the real-time and simultaneous observation of
neuron activities from thousands of neurons, producing individual time series
representing their fluorescence intensity. From these data, the connectome
inference problem amounts to retrieving the synaptic connections between
neurons on the basis of the fluorescence time series. This
problem is often difficult because of experimental issues, including
masking effects (i.e., some of the neurons are not observed or confounded with others), the low sampling rate of the optical device with respect to the
neural activity speed or the slow decay of fluorescence.

@@

In this paper, we present a simplified -- yet nearly as good -- version of the
winning method\footnote{Code available at \url{https://github.com/asutera
/kaggle-connectomics}} of the Connectomics
Challenge\footnote{\url{http://connectomics.chalearn.org}}, as a simple and
theoretically grounded approach based on signal processing techniques and
partial correlation statistics. The paper is structured as follows:
Section~\ref{sec:filter} describes the signal processing methods applied on
fluorescent calcium time series; Section\ref{sec:inference} then presents the
proposed approach and its theoretical properties; Section~\ref{sec:results}
provides an empirical analysis and comparison with other network inference
methods, while Section~\ref{sec:conclusion} finally discusses our work and
provides further research directions. Additionally,
Appendix~\ref{app:optimized} further describes, in all details, our actual
winning method, giving slightly better results than the method presented in
the paper, at the price of parameter tuning. Appendix~\ref{app:supp} provides supplementary results on other datasets.

@@

Under the simplifying assumption that neurons are on-off units, characterized
by short periods of intense activities, or peaks, and longer periods of
inactivity, the first part of our algorithm consists in cleaning the raw
fluorescence data. More specifically, time series are processed using standard
signal processing filters in order to remove noise due to light scattering
effects, to account for fluorescence  low decay and to reduce the importance of
global high activity in the network. The overall process is illustrated in
Figure~\ref{fig:filtered-signal}.

@@

As Figure~\ref{fig:original_curve} shows, the raw
fluorescence signal is very noisy due to light scattering artifacts that
usually affect the quality of the recording~\citep{lichtman2011big}.
Accordingly, the first step of our pipeline is to smoothen the signal, using
one of the following low-pass filters for filtering out high frequency noise:
\begin{align}
% symmetrical median filter
f_1(x^t_i) &= x^{t-1}_i + x^t_i + x^{t+1}_i, \label{eq:symetric-median} \\
% asymmetrical weighted median
f_2(x^t_i) &= 0.4 x^{t-3}_i + 0.6 x^{t-2}_i + 0.8 x^{t-1}_i + x_i^t.
\label{eq:weighted-asymetric-median}
\end{align}
For the sake of illustration, the effect of the filter $f_1$ on the signal
is shown in Figure \ref{fig:lp_curve}.

Furthermore, it can be observed that neurons communicate through short spikes, characterized by a high
frequency, while low frequencies of the signal mainly correspond to the slow
decay of fluorescence. To have a signal that only has high magnitude around instances where the spikes occur, the second step of our pipeline transforms the time series into its backward
difference
\begin{align}
g(x^{t}_{i}) &= x^{t}_i - x^{t-1}_i, \label{eq:high-pass-filter}
\end{align}
as shown in Figure \ref{fig:hp_curve}.

To filter out small variations in the signal obtained after applying the
function $g$ as well as to eliminate negative values, we use the following
hard-threshold filter
\begin{align}\label{eqn:hfilter}
h(x^{t}_i) &= x^{t}_i \mathbb{1}(x^{t}_i \geq \tau) \text{ with } \tau > 0,
\end{align}
yielding Figure \ref{fig:threshold_curve}.
As we can see, the processed signal only contains clean spikes.

@@

Practically, the computation of all $p_{i,j}$ coefficients using Equation
\ref{eq:inverse} requires the estimation of the covariance matrix $\Sigma$
and then computing its inverse. Given that we have typically more
samples than neurons, the covariance matrix can be inverted in a
straightforward way. We nevertheless obtained some improvement by
replacing the exact inverse with an approximation using only the $M$
first principal components \citep{bishop2006pattern} (with
$M=0.8 p$ in our experiments).

@@

% Note on directionality

%% Our  connection inference  procedure  relies  on several  assumptions. First,
%% we will assume that  the fluorescence concentrations of all $p$ neurons can be
%% modeled as a set of random variables $X = \{X_1, \dots, X_p\}$  that are drawn
%% from a time invariant joint  probability distribution $P_X$. Or, say otherwise,
%% we assume that there exists a probability distribution $P_X$ such that the
%% fluorescence concentrations at a particular time $t$ can be seen as an
%% independent realization of $P_X$.

%% To assess the presence of a connection between two neurons, we will
%% use partial correlation statistics, which has been used extensively
%% for the inference of gene regulatory networks \cite{}.

%% To a joint probability distribution $P_X$ corresponds a
%% graph where (i) each node is a random variable and (ii) two nodes are
%% only connected if they correspond to two dependent variables. We assume that two neurons are
%% connected if  and only if,  there exists  in this  graph an edge
%% that connects  directly the  two random  variables to  which they correspond.
%% In this context, it is equivalent to writing that two neurons $i$ and $j$ are  not  directly
%% connected  if and only if  $X_i$  and $X_j$  are  conditionally independent
%% with  respect to all  other variables. We remind  that two random  variables
%% $X_i$  and  $X_j$  are  said  to  be  conditionally independent to a (possibly
%% empty)  set of controlling random variables $Z$, denoted $X_i \perp X_j | Z$,
%% if the joint conditional probability distribution   $P_{X_i,  X_j   |   Z}$
%% factorizes  into   $P_{X_i|Z} P_{X_j|Z}$.  In particular,  when  $X_i$ and
%% $X_j$ are  conditionally independent  with respect  to all  the other
%% variables, denoted  $X_i \perp  X_j  | X^{-i,j}$,  then  the  observation  of
%% $X_i$  given  the knowledge  of  $X^{-i,j}$ provides  no information  about
%% $X_j$,  and vice-versa.

%% In the  particular case where $P_X$ is  a joint  Gaussian distribution
%% $\mathcal{N}(\mu; \Sigma)$ of mean  vector $\mu$ and covariance matrix
%% $\Sigma$, $X_i$  and $X_j$ are conditionally  independent with respect to the
%% other variables $X^{-i,j}$ if their \textit{partial correlation} statistic is
%% null~\citep{baba2004partial}, i.e., if
%% \begin{equation}
%% \rho_{X_i, X_j | X^{-i,j}} = -\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}} = 0, \label{eq:inverse}
%% \end{equation}
%% where $\Sigma^{-1}$ is known as the precision or concentration matrix.
%% Conversely, $X_i$ and $X_j$ are conditionally dependent if $\rho_{X_i, X_j |
%% X^{-i,j}} \neq 0$, where $\rho_{X_i, X_j | X^{-i,j}} \in [-1;1]$. Note that the
%% larger the value of this term, the stronger the dependence between variables
%% $X_i$ and $X_j$.

%% Our approach implicitly makes the additional assumption  that $P_X$ is  a
%% joint Gaussian distribution since it  outputs as degree of association
%% between two neurons $i$ and $j$ an estimate of $\rho_{X_i, X_j | X^{-i,j}}$.
%% Note also that the partial correlation
%% statistic is symmetric  (i.e., $\rho_{X_i, X_j | X^{-i,j}} = \rho_{X_j, X_i |
%% X^{-j,i}}$). Therefore, our approach cannot identify the direction of the interaction. We will see in Section~\ref{sec:results} why this only affects slightly its performances, with respect to the metric used in the Connectomics challenge.

%% We now detail our procedure for estimating partial correlation
%% statistics ($\rho_{X_i, X_j | X^{-i,j}}$) from the data. It first
%% estimates the covariance matrix from the data. Afterwards, it computes
%% the exact inverse of this matrix, from which it infers the value of
%% $\rho_{X_i, X_j | X^{-i,j}}$ using equation (\ref{eq:inverse}). Note
%% that we have also experimented procedures where the exact inverse was
%% replaced by an approximation of it using only the $M$ principal
%% components of $\Sigma$~\citep{bishop2006pattern} whose associated
%% eigen values were significantly higher than zero. Such an
%% approximation is equivalent to assume that the firing activity of all
%% neurons can be explained by only a subset of them whose size is equal
%% to $M$. We observed that those procedures were often leading to better
%% results. Note that we found out by running expriments that the number
%% $M$ of principal components is around $0.8p$ (see discussion in
%% Appendix~\ref{app:REF} \textcolor{red}{REF}).

%% % Computationally, they can be estimated
%% % % efficiently in $O(M^3)$ (instead of $O(p^3)$)
%% % from the $M$ first principal
%% % components extracted through PCA~\citep{bishop2006pattern} on the processed
%% % data. While speeding up computation, it allows to further
%% % filter the the data by removing the contribution of components that are respectively
%% % due to redundancy or noise, i.e., with an eigen value respectively is equal to zero or close to zero.

@@

%% \textcolor{red}{to incorporate:
%% To highlight the benefits of measuring the partial dependence conditional to
%% all other variables, we compare in Section~\ref{sec:results} with the
%% \textit{Pearson correlation} statistics, which measure instead the linear
%% dependence between variables, independently of the others. Accordingly, $X_i$
%% and $X_j$ are independent, without any conditioning, if their Pearson correlation statistic is zero, that
%% is if
%% \begin{equation}
%% \rho_{X_i,X_j} = \frac{\Sigma_{ij}}{\sqrt{\Sigma_{ii} \Sigma_{jj}}} = 0.
%% \end{equation}}

@@

% \begin{table}[t]\label{tab:comparison}
% %\caption{Evaluation of different methods on the four \textit{normal} datasets\label{tab:comparison}}
% \centering
% \small
% \begin{tabular}{| l | c c c c c c | c c c c c c|}
% \hline
% & \multicolumn{6}{c|}{AUROC} & \multicolumn{6}{c|}{AUPRC} \\
% \textit{Method} $\backslash$ \textit{normal-} & \textit{lowcon} & \textit{highcon} & \textit{lowcc} & \textit{highcc} & \textit{3-highrate} & \textit{4-lownoise} & \textit{lowcon} & \textit{highcon} & \textit{lowcc} & \textit{highcc} & \textit{3-highrate} & \textit{4-lownoise}\\
% \hline
% \hline
% % No  filtering       					& 0.777 & 0.767 & 0.772 & 0.774 & 0.070 & 0.064 & 0.068 & 0.072\\
% % $ h \circ g \circ f_1$                  & 0.923 & 0.925 & 0.923 & 0.922 & 0.311 & 0.315 & 0.313 & 0.304\\
% % $ w \circ h \circ g \circ f_1$          & 0.931 & 0.929 & 0.928 & 0.926 & 0.326 & 0.323 & 0.319 & 0.303\\
% % + PCA         							& 0.932 & 0.930 & 0.928 & 0.926 & 0.355 & 0.353 & 0.350 & 0.333\\
% % Averaging           					& 0.937 & 0.935 & 0.935 & 0.931 & 0.391 &  0.390 &  0.385 & 0.375\\
% % Full method           					& 0.943 & 0.942 & 0.942 & 0.939 & 0.403 & 0.404 & 0.398 & 0.388\\
% % \hline
% % PC & 0.886 & 0.884 & 0.891 &  0.877 & 0.153 & 0.145 & 0.170 & 0.132\\
% % GTE & 0.890 & 0.893 & 0.894 & 0.873 & 0.171 & 0.174 & 0.197 & 0.142\\
% % GENIE3 & & & & & & & & \\
% % \hline
% \end{tabular}
% \caption{Top: Performance on \textit{normal-1,2,3,4} with partial correlation and different filtering functions.
% Bottom: Performance on \textit{normal-1,2,3,4} with different methods.}
% \end{table}

% NEXT APPENDIX SECTION
% SUPPLEMENTARY RESULTS
% EXPLAIN EACH TAB AND VERIFY THE EXACTITUDE - RUN OPTIMIZED VERSION ON low-noise, ... ?
%

% EXPLAIN PCA - FIG

%\vspace{5cm}

% The algorithm propose in sections \ref{sec:filter} and \ref{sec:inference} has been optimized at its extreme limit to win the challenge. In this section, we will describe the full tuned algorithm and present some results.

% Note that our parameter tuning was mainly
% done on the \textit{normal-1} dataset and partly on \textit{normal-4} dataset,
% therefore it may justify why these networks are better retrieved despite the
% fact that some might be simpler to infer.

% % Introduce a custom solution to make the Y matrix asymmetric
% % Introduce directivity
% \textcolor{red}{Est-ce qu'on ne le ferait pas passer dans l'appendice B? Ou du moins l'équation?}
% Since partial correlation is a symmetric measure, the causal mechanism behind the
% interaction ($A$ directly causes $B$, $B$ directly causes $A$ or both) can not
% be directly inferred\footnote{Also note that two other causal mechanisms might be
% implied while having a non-zero partial coefficient: (i) a pair of variables
% is induced by a common (hidden) variable; (ii) $A$ (respectively $B$) is
% conditionally correlated to a hidden variable affecting $B$ (resp. $A$)
% \citep{de2004discovery}.}.
% In order to retrieve causal information, we developed an
% heuristics based on the following observation. The rise of fluorescence
% indicates the activation of a neuron $i$. If a neuron $j$ have also
% an increased of fluorescence after a slight delay, this could have been
% caused by the neuron $i$. Thus, we count whenever
% such events occurred between consecutive time steps
% \[
% \hat{Z}: z_{ij} = \sum_{t=1}^{T - 1}
%     \mathbb{1}((x_j^{t+1} - x_i^t) \in \left[f_1, f_2 \right]) -
%     \mathbb{1}((x_i^{t+1} - x_j^t) \in \left[f_1, f_2 \right]), \forall i, j \in V
% \]
% where $f_1$ and $f_2$ are parameters of the method.
% Finally, the weighted average of correlation matrices is combined with $Z$ such
% that $Z$ has a contribution of $0.3\%$ to the prediction of an edge.

% % For the directory method, only filters defined by Equations
% % \ref{eq:low-pass}, \ref{eq:high-pass} and \ref{eq:hard-treshold-filter} were
% % used.

% \section{Extended results}

% \begin{table}
% \small
% \centering
% \begin{tabular}{@{}l *{8}{c}@{}}
%   & \multicolumn{4}{c}{AUROC} & \multicolumn{4}{c}{AUPRC} \\
% \hline \hline
% Methods $\backslash$ normal- & 1 & 2 & 3 & 4 & 1 & 2 & 3 & 4 \\
% \hline \hline
% % Directivity \textcolor{red}{appendix?} & 0.535 & 0.526 & 0.533 & 0.535 & 0.012 & 0.012 & 0.012 & 0.012 \\
% Pearson correlation    & 0.886 & 0.884 & 0.891 & 0.877 & 0.153 & 0.145 & 0.170 & 0.132 \\
% GENIE3 (depth = 1)     & 0.878 & 0.877 & 0.873 & 0.873 & 0.184 & 0.173 & 0.182& 0.172\\
% GENIE3 (depth = 2)     & 0.889 & 0.888 & 0.883 & 0.884* & 0.217 & 0.210 & 0.220 & 0.199* \\
% GENIE3 (depth = 3)     & 0.892 & 0.891* & 0.887 & 0.887 & 0.232 & 0.226* & 0.237 & 0.215\\
% GTE                    & 0.890 & 0.893 & 0.894 & 0.873 & 0.171 & 0.174 & 0.197 & 0.142 \\
% Partial correlation    & 0.930 &  0.927 &  0.926 &  0.923& 0.363  & 0.347 &  0.346 & 0.328 \\
% Partial correlation (avg. 2) & 0.937 & 0.935 & 0.935 & 0.931 & 0.391 &  0.390 &  0.385 & 0.375  \\
% Partial correlation (avg. 4) & 0.938 & 0.936 & 0.936 & 0.932& 0.391 & 0.389 & 0.385 & 0.374\\
% % Our approach simple + dir & 0.939 & 0.936 & 0.937 & 0.933& 0.392 & 0.390 & 0.386 & 0.375\\
% Partial correlation (opt.) & 0.932 & 0.931 & 0.930 & 0.928 & 0.364 & 0.366 & 0.359 & 0.344 \\
% Our approach (opt. \& avg. 4)    & 0.943 & 0.942 & 0.942 & 0.939 & 0.403 & 0.404 & 0.398 & 0.388 \\
% % Our approach + dir  \textcolor{red}{appendix?}   & 0.944 & 0.943 & 0.942 & 0.940 & 0.404 & 0.405 & 0.399 & 0.389 \\
% \end{tabular}
% \caption{TODO (* means not 10000 trees but more than 9500, to be solved.) }
% \label{tab:tab1}
% \end{table}

% \section{Supplementary results}

% \begin{table}[htbp]
% \centering
% \caption{Correlation (ROC). Improvement of each stage of our approach. Note that we choose the
%          \textit{best} threshold (\textcolor{red}{for now 0.11}) for stages 1 to 4.}
% \begin{tabular}{*{5}{l}}
% \toprule
% Stage               & normal-1 & normal-2 & normal-3 & normal-4 \\
% \midrule
% Nothing             & 0.681 & 0.699 & 0.683 & 0.681\\
% Weights             & 0.695 & 0.713 & 0.697 & 0.694\\
% H-T                 & 0.876 & 0.873 & 0.877 & 0.869\\
% H-T + Weights       & 0.886 & 0.884 & 0.891 & 0.877\\
% P-P                 & 0.857 & 0.850 & 0.843 & 0.832\\
% P-P + Weights       & 0.826 & 0.823 & 0.817 & 0.799\\
% \bottomrule
% \end{tabular}
% \end{table}

% \begin{table}[htbp]
% \centering
% \caption{Correlation (P-R). Improvement of each stage of our approach. Note that we choose the
%          \textit{best} threshold (\textcolor{red}{for now 0.11}) for stages 1 to 4.}
% \begin{tabular}{*{5}{l}}
% \toprule
% Stage               & normal-1 & normal-2 & normal-3 & normal-4 \\
% \midrule
% Nothing             & 0.028 & 0.028 & 0.025 & 0.026\\
% Weights             & 0.031 & 0.030 & 0.027 & 0.028\\
% H-T                 & 0.143 & 0.128 & 0.150 & 0.129\\
% H-T + Weights       & 0.153 & 0.145 & 0.170 & 0.132\\
% P-P                 & 0.100 & 0.087 & 0.079 & 0.083\\
% P-P + Weights       & 0.079 & 0.071 & 0.067 & 0.064\\
% \bottomrule
% \end{tabular}
% \end{table}

% \begin{table}[htb]
% \centering
% \caption{Comparison (P-R) with other methods (for now, $t = 0.11$)}
% \begin{tabular}{*{5}{l}}
% \toprule
% Methods             & normal-1 & normal-2 & normal-3 & normal-4 \\
% \midrule
% Correlation (best)         & 0.153 & 0.145 & 0.170 & 0.132 \\
% Partial correlation (best) & 0.364 & 0.366 & 0.359 & 0.344 \\
% Our approach (best)        & 0.403 & 0.404 & 0.398 & 0.388 \\
% GENIE3                     & & & & \\
% GTE                        & 0.171 & 0.174 & 0.197 & 0.142 \\
% Directivity                & 0.012 & 0.012 & 0.012 & 0.012 \\
% Our approach + dir         & 0.404 & 0.405 & 0.399 & 0.389 \\
% \bottomrule
% \end{tabular}
% \end{table}

% \begin{table}[htb]
% \centering
% \caption{Inverse correlation (ROC). Improvement of each stage of our approach. Note that we choose the
%          \textit{best} threshold (\textcolor{red}{for now 0.11}) for stages 1 to 4.}
% \begin{tabular}{*{5}{l}}
% \toprule
% Stage               & normal-1 & normal-2 & normal-3 & normal-4 \\
% \midrule
% Nothing             & 0.777 & 0.767 & 0.772 & 0.774 \\
% PCA                 & 0.780 & 0.770 & 0.776 & 0.777 \\
% Weights             & 0.780 & 0.769 & 0.776 & 0.777 \\
% Weights + PCA       & 0.783 & 0.772 & 0.779 & 0.780 \\
% H-T                 & 0.893 & 0.891 & 0.891 & 0.886 \\
% H-T + PCA           & 0.894 & 0.892 & 0.891 & 0.886 \\
% H-T + Weights       & 0.899 & 0.897 & 0.896 & 0.891 \\
% H-T + Weights + PCA & 0.900 & 0.898 & 0.896 & 0.892 \\
% P-P                 & 0.925 & 0.925 & 0.924 & 0.923 \\
% P-P + PCA           & 0.926 & 0.926 & 0.925 & 0.923 \\
% P-P + Weights       & 0.931 & 0.930 & 0.928 & 0.927 \\
% P-P + Weights + PCA & 0.932 & 0.931 & 0.930 & 0.928 \\
% Averaging (PCA)     & 0.943 & 0.942 & 0.942 & 0.939 \\
% Stacking            & 0.944 & 0.943 & 0.942 & 0.940 \\
% \bottomrule
% \end{tabular}
% \end{table}

>>>>

\vspace{-0.8cm}
\begin{abstract}
In this work, we propose a simple, but yet efficient method for the problem of
connectome inference in calcium imaging data. The proposed algorithm consists of
two steps. First, processing the raw signals to detect neural peak activities.
Second, inferring the degree of association between neurons from partial
correlation statistics. This paper summarizes the methodology that led us to
win the Connectomics Challenge, proposes a simplified version of our method and
finally discusses our results with respect to other inference methods.
\end{abstract}

@@

The human brain is a complex biological
organ made of about 100 billions of neurons, each connected to 7000 other neurons on
average \citep{pakkenberg2003aging}. Unfortunately, direct observation of the connectome, the wiring
diagram of the brain, is not yet technically feasible. Without being perfect,
calcium imaging currently allows the real-time and simultaneous observation of
neuron activities from thousands of neurons, producing individual time series
representing their fluorescence intensity. From these data, the connectome
inference problem amounts to retrieving the synaptic connections between
neurons on the basis of the fluorescence time series. This
problem is often difficult because of experimental issues, including
masking effects (i.e., some of the neurons are not observed or confounded with others), the low sampling rate of the optical device with respect to the
neural activity speed or the slow decay of fluorescence.

@@

In this paper, we present a simplified -- yet nearly as good -- version of the
winning method\footnote{Code available at \url{https://github.com/asutera/kaggle-connectomics}} of the Connectomics
Challenge\footnote{\url{http://connectomics.chalearn.org}}, as a simple and
theoretically grounded approach based on signal processing techniques and
partial correlation statistics. The paper is structured as follows:
Section~\ref{sec:filter} describes the signal processing methods applied on
fluorescent calcium time series; Section\ref{sec:inference} then presents the
proposed approach and its theoretical properties; Section~\ref{sec:results}
provides an empirical analysis and comparison with other network inference
methods, while Section~\ref{sec:conclusion} finally discusses our work and
provides further research directions. Additionally,
Appendix~\ref{app:optimized} further describes, in all details, our actual
winning method, giving slightly better results than the method presented in
the paper, at the price of parameter tuning. Appendix~\ref{app:supp} provides supplementary results on other datasets.

@@

Under the simplifying assumption that neurons are on-off units, characterized
by short periods of intense activities, or peaks, and longer periods of
inactivity, the first part of our algorithm consists in cleaning the raw
fluorescence data.
More specifically, time series are processed using standard
signal processing filters in order to (i) remove noise mainly due to fluctuations independent of calcium, calcium fluctuations independent of spiking activity, calcium fluctuations in nearby tissues that have been mistakenly captured, or simply by the imaging process (ii) to account for fluorescence low decay and (iii) to reduce the importance of
global high activity in the network. The overall process is illustrated in
Figure~\ref{fig:filtered-signal}.

@@

As Figure~\ref{fig:original_curve} shows, the raw
fluorescence signal is very noisy due to light scattering artifacts that
usually affect the quality of the recording~\citep{lichtman2011big}.
Accordingly, the first step of our pipeline is to smoothen the signal, using
one of the following low-pass filters for filtering out high frequency noise:
\begin{align}
% symmetrical median filter
f_1(x^t_i) &= x^{t-1}_i + x^t_i + x^{t+1}_i, \label{eq:symetric-median} \\
% asymmetrical weighted median
f_2(x^t_i) &= 0.4 x^{t-3}_i + 0.6 x^{t-2}_i + 0.8 x^{t-1}_i + x_i^t.
\label{eq:weighted-asymetric-median}
\end{align}
These filters are standard in the signal processing field \citep{kaiser1977data, oppenheim1983signals}. The idea of the second one is to not consider future values and to introduce decreasing weights. For the sake of illustration, the effect of the filter $f_1$ on the signal
is shown in Figure \ref{fig:lp_curve}.

Furthermore, short spikes, characterized by a high
frequency, can be seen as an indirect indicator of neuron communications while low frequencies of the signal mainly correspond to the slow
decay of fluorescence. To have a signal that only has high magnitude around instances where the spikes occur, the second step of our pipeline transforms the time series into its backward
difference
\begin{align}
g(x^{t}_{i}) &= x^{t}_i - x^{t-1}_i, \label{eq:high-pass-filter}
\end{align}
as shown in Figure \ref{fig:hp_curve}.

To filter out small variations in the signal obtained after applying the
function $g$ as well as to eliminate negative values, we use the following
hard-threshold filter
\begin{align}\label{eqn:hfilter}
h(x^{t}_i) &= x^{t}_i \mathbb{1}(x^{t}_i \geq \tau) \text{ with } \tau > 0,
\end{align}
yielding Figure \ref{fig:threshold_curve} where $\tau$ is the threshold parameter and $\mathbb{1}$ is the indicator function.
As we can see, the processed signal only contains clean spikes.

@@

Practically, the computation of all $p_{i,j}$ coefficients using Equation
\ref{eq:inverse} requires the estimation of the covariance matrix $\Sigma$
and then computing its inverse. Given that we have typically more
samples than neurons, the covariance matrix can be inverted in a
straightforward way. We nevertheless obtained some improvement by
replacing the exact inverse with an approximation using only the $M$
first principal components \citep{bishop2006pattern} (with
$M=0.8 p$ in our experiments, see Appendix~\ref{app:pca}).

@@

@@

@@

\section{About the number of principal components}
\label{app:pca}

The main idea of principal component analysis (PCA) is to reduce the
dimensionality of a dataset whose variables are interrelated
\citep{jolliffe2005principal}. It consists in finding a new ordered set of
uncorrelated variables, called the principal components, where the most
principal components, i.e. the first variables, hold most of the data variation, also denoted as
variance, scattered in all the original variables.

Figure~\ref{fig:pca} shows the explained variance by compoments and gives a hint about the number of uncorrelated variables required to
explain all the variation in the original data.

The singular value decomposition (SVD) relies on the fact that every matrix -
not only square matrices - can be decomposed into a diagonal nonnegative
matrix with the help of two unit matrices \citep{alter2000singular}.

The terms of the diagonal matrix can be used to determine the fraction of the
overall information hold by the components in the decomposition and therefore
give the relative significance of the components \citep{alter2000singular}.
Figure~\ref{fig:SVD} displays the singular value ratio of all networks in the
challenge.

Obvioulsy, Figure~\ref{fig:pca} shows that for both approaches only 800
components are required to explain the (majority of) network structures.
Therefore, the value of the parameter $M$ is immediat and should be obviously
set to $800$, or equivalently to $0.8p$. \\

\begin{figure}[t]
\centering
\subfigure[PCA]{\includegraphics[width=0.45\textwidth]{explained_variance_all} \label{fig:PCA}}
\subfigure[SVD]{\includegraphics[width=0.45\textwidth]{singular_values_all} \label{fig:SVD}}
\caption{Explained variance ratio by component (left) and singular value ratio by component (right) for all networks.}
\label{fig:pca}
\end{figure}
\section{Computational information}

\begin{table}[h]
\caption{Execution time, scalability, memory requirements for methods described in the paper. \textcolor{red}{to do.}}
\label{tab:results_appendix}
\centering
\small
\begin{tabular}{| l|c|c|c|} \hline
Methods & Execution time & Scalability & Memory requirements \\ \hline
Partial correlation & & & \\ 
GTE  & & & \\
GENIE3 & & & \\ \hline
\end{tabular}
\end{table}
